---
title: "Titanic Dataset Project"
author: "Anfernee Charles, Shannon-Rae Hackett, Leeza, Vishal"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r reading data}
library(tidyverse)
library(naniar)
library(explore)
library(DataExplorer)
library(dplyr)
library(simputation)
library(ppsr) 
library(correlationfunnel)
library(tidymodels)
library(xgboost)
library(vip)
library(randomForest)
library(readxl)
library(rpart)


```

```{r}
library(cowplot)
library(mice)


train <- read_excel("C:/Users/anfer/OneDrive/Desktop/Module 4/FinalProject/CSV/CSV/train.xlsx")

test <- read_excel("C:/Users/anfer/OneDrive/Desktop/Module 4/FinalProject/CSV/CSV/test.xlsx")

df_train <- train
df_test <- test

df_train |> head(10)

```
```{r}
df_test |> head(10)
```



# Data dictionary:

## Variable Definition Key

-   **Survival**, Survival 0 = No, 1 = Yes
-   **Pclass**, Ticket class 1 = 1st, 2 = 2nd, 3 = 3rd
-   **Sex**, Sex
-   **Age**, Age in years
-   **Sibsp**, \# of siblings / spouses aboard the Titanic
-   **Parch**, \# of parents / children aboard the Titanic
-   **Ticket**, Ticket number
-   **Fare**, Passenger fare
-   **Cabin**, Cabin number
-   **Embarked**, Port of Embarkation C = Cherbourg, Q = Queenstown, S = Southampton

```{r}
df_train |> describe() 
```

```{r}
df_test |> describe()
```


```{r}
## Renaming variables for better readability

 df_train <- df_train |>
  rename("siblings_spouses_aboard" = "SibSp",
         "parents_children_aboard" = "Parch",
         "port_of_embarkment" = "Embarked")
```

```{r}
# doing the same on the test data

df_test <- df_test |>
  rename("siblings_spouses_aboard" = "SibSp",
         "parents_children_aboard" = "Parch",
         "port_of_embarkment" = "Embarked")
```


```{r}
## changing character dtypes into factor

df_train$Name <- as.factor(df_train$Name)

df_train$Sex <- as.factor(df_train$Sex)

df_train$Ticket <- as.factor(df_train$Ticket)

df_train$Cabin <- as.factor(df_train$Cabin)

df_train$port_of_embarkment <- as.factor(df_train$port_of_embarkment)



df_test$Name <- as.factor(df_test$Name)

df_test$Sex <- as.factor(df_test$Sex)

df_test$Ticket <- as.factor(df_test$Ticket)

df_test$Cabin <- as.factor(df_test$Cabin)

df_test$port_of_embarkment <- as.factor(df_test$port_of_embarkment)





```

```{r}

## Analyzing the dataset

plot_intro(df_train)
```


```{r}
plot_intro(df_test)
```


```{r missing values}
## Viewing missing values

vis_miss(df_train)
```

```{r}
vis_miss(df_test)
```

```{r}
visdat::vis_dat(df_train)
```


```{r}
visdat::vis_dat(df_test)
```

# Missing values

## 8.1% of the total training data set contain missing values, and 9% of the total test data set contain missing values. Cabin has 77% missing values, with Age having 20%.

```{r}
# dropping Name, Cabin, PassengerID and Ticket because they'll have no influence on the prediction of the target variable.

df_train <- df_train |> select(-Cabin, -Name, -Ticket, -PassengerId)

df_test <- df_test |> select(-Cabin, -Name, -Ticket, -PassengerId)

```

```{r}
plot_density(df_train)
```

```{r focusing on Age}

plot_density(df_train$Age)

```



# Exploring different methods of imputation

```{r using simputation on Age}

# Iteration #1

Age_imp <- df_train |>
  bind_shadow() |>
  impute_lm(Age ~ Pclass) |>
  add_label_shadow()

ggplot(Age_imp,
       aes(x = Age,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")
```

```{r}

#Iteration #2
# adding all other features to the imputation

Age_imp <- df_train |>
  bind_shadow() |>
  impute_lm(Age ~ Pclass + Sex + Fare + port_of_embarkment + siblings_spouses_aboard) |>
  add_label_shadow()

ggplot(Age_imp,
       aes(x = Age,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")

```

```{r }
#Iteration #3

# creating a data frame with imputed values and creating a plot that compares it to the original data


Age_imp <- data.frame(
  original = df_train$Age,
  imputed_zero = replace(df_train$Age, is.na(df_train$Age), 0),
  imputed_mean = replace(df_train$Age, is.na(df_train$Age), mean(df_train$Age, na.rm = TRUE)),
  imputed_median = replace(df_train$Age, is.na(df_train$Age), median(df_train$Age, na.rm = TRUE))
)




h1 <- ggplot(Age_imp, aes(x = original)) +
  geom_histogram(fill = "tomato", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Original distribution") +
  theme_classic()

h2 <- ggplot(Age_imp, aes(x = imputed_zero)) +
  geom_histogram(fill = "green", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Zero-imputed distribution") +
  theme_classic()

h3 <- ggplot(Age_imp, aes(x = imputed_mean)) +
  geom_histogram(fill = "royalblue", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Mean-imputed distribution") +
  theme_classic()

h4 <- ggplot(Age_imp, aes(x = imputed_median)) +
  geom_histogram(fill = "gold", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Median-imputed distribution") +
  theme_classic()

plot_grid(h1, h2, h3, h4, nrow = 2, ncol = 2)

```

## All of the above distributions were severely impacted when compared to the original distribution.

```{r using MICE}
# iteration #4 using Multivariate Imputation via Chained Equations (MICE)

numeric_cols <- df_train |>
  select(Pclass, siblings_spouses_aboard, parents_children_aboard, Age)
```

```{r}
mice_imp <- data.frame(
  original = df_train$Age,
  pmm_imp = complete(mice(numeric_cols, method = "pmm"))$Age,
  cart_imp = complete(mice(numeric_cols, method = "cart"))$Age,
  lasso_imp = complete(mice(numeric_cols, method = "lasso.norm"))$Age
  
)


```

```{r}

mh1 <- ggplot(mice_imp, aes(x = original)) +
  geom_histogram(fill = "tomato", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Original distribution") +
  theme_classic()

mh2 <- ggplot(mice_imp, aes(x = pmm_imp)) +
  geom_histogram(fill = "green", color = "#000000", position = "identity", bins = 50) +
  ggtitle("PMM-imputed distribution") +
  theme_classic()

mh3 <- ggplot(mice_imp, aes(x = cart_imp)) +
  geom_histogram(fill = "royalblue", color = "#000000", position = "identity", bins = 50) +
  ggtitle("CART-imputed distribution") +
  theme_classic()

mh4 <- ggplot(mice_imp, aes(x = lasso_imp)) +
  geom_histogram(fill = "gold", color = "#000000", position = "identity", bins = 50) +
  ggtitle("Lasso-imputed distribution") +
  theme_classic()


plot_grid(mh1, mh2, mh3, mh4, nrow = 2, ncol = 2)
```

# CART imputed Age distribution looks the closest to the original distribution

```{r}
#replacing values in age with imputed values from CART imputation 

df_train$Age[is.na(df_train$Age)] <- mice_imp$cart_imp[is.na(df_train$Age)]

df_test$Age[is.na(df_test$Age)] <- mice_imp$cart_imp[is.na(df_test$Age)]
```


```{r}
# verifying change in distribution

train_plt <- ggplot(df_train, aes(x = Age)) +
  geom_histogram(fill = "gold", color = "#000000", position = "identity", bins = 50) +
  ggtitle("imputed distribution") +
  theme_classic()


plot_grid(mh1, train_plt, ncol = 2, nrow = 1)

```

## Checking further missing values

```{r}
df_train |> describe()
```

```{r}
df_test |> describe()
```



```{r}
# dropping all missing rows

df_train <- df_train |> drop_na()


df_test <- df_test |> drop_na()

```



## Feature Creation

```{r}
# creating a column to distinguish between child and adult

df_train <- df_train |> mutate(Child = if_else(df_train$Age >= 18, 'Adult','Child'))


df_test <- df_test |> mutate(Child = if_else(df_test$Age >= 18, 'Adult','Child'))

```

```{r}
# viewing how many children and adults survived

table(df_train$Child, df_train$Survived)
```

## 474 and 75 Adults and children died respectively, while 277 and 65 Adults and Children survived, respectively.

```{r}
# changing Child column to factor

df_train$Child <- as.factor(df_train$Child)

df_test$Child <- as.factor(df_test$Child)


```

```{r}
# creating a plot showing amount of children and adults that survived

ggplot(data=df_train, aes(Child, fill = factor(Survived))) + 
  geom_bar(stat = "count", position = "dodge") + 
  xlab("Adults and Children") + 
  scale_fill_discrete(name = "Survived") + 
  ggtitle("Survivors by Number of Adults or Children")
```

```{r}
# creating a relatives aboard variable

df_train <- df_train |> mutate(Relatives_aboard = df_train$siblings_spouses_aboard + df_train$parents_children_aboard)

df_test <- df_test |> mutate(Relatives_aboard = df_test$siblings_spouses_aboard + df_test$parents_children_aboard)

```

```{r}
# discretizing the family size variable

df_train$Family_size[df_train$Relatives_aboard == 1] <- 'Single'
df_train$Family_size[df_train$Relatives_aboard < 5 & df_train$Relatives_aboard > 1] <- 'Small'
df_train$Family_size[df_train$Relatives_aboard > 4] <- 'Large'
df_train$Family_size[df_train$Relatives_aboard == 0] <- 'None'



df_test$Family_size[df_test$Relatives_aboard == 1] <- 'Single'
df_test$Family_size[df_test$Relatives_aboard < 5 & df_test$Relatives_aboard > 1] <- 'Small'
df_test$Family_size[df_test$Relatives_aboard > 4] <- 'Large'
df_test$Family_size[df_test$Relatives_aboard == 0] <- 'None'



# changing data type to factor

df_train$Family_size <- as.factor(df_train$Family_size)

df_test$Family_size <- as.factor(df_test$Family_size)
```
```{r}
# creating a plot of family size to view which category had the highest survival count

ggplot(data=df_train, aes(Family_size, fill = factor(Survived))) + 
  geom_bar(stat = "count", position = "dodge") + 
  xlab("Family size") + 
  scale_fill_discrete(name = "Survived") + 
  ggtitle("Survivors by Family size")
```


```{r}
# looking at the relationship of all the categorical features with survived

plot_bar(df_train, by = "Survived")
```





### * Women were more likely to survive then men

### * persons from port of embarkation C (Cherbourg) had a higher survival rate

### * more children (< 18 y/o) survived than adults

### * Single and small (<4) family sizes were more likely to survive









# Correlation plots and feature importance
## * generating the pp score





```{r}
survived_ppsr_score <- df_train |>
    score_predictors(y = 'Survived', do_parallel = TRUE) |>
    as_tibble()

survived_ppsr_score |> glimpse()

```
```{r}
# plotting 1bar pps (predictive power score)

df_train |> visualize_pps(y = "Survived", do_parallel = TRUE)


```






### * Sex has the highest pp score, while relatives_aboard, family size, Pclass and Fare have lower pp scores






```{r}

#plotting a pps matrix to expand on feature relationships

g <- df_train |>
    visualize_pps(do_parallel = TRUE)

g +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
# utilizing a correlation funnel which is a binary version of the correlation matrix

survived_binned_tbl <- df_train |>
    binarize()

#survived_binned_tbl |> glimpse()

survived_binned_tbl |>
    correlate(target = Survived__1) |>
    plot_correlation_funnel() +
    geom_point(size = 3, color = "black")


```




### * Sex = Female, Pclass = 1, Fare > $31, Family_size = small and port_of_embarkment = C are all positively related to survived = 1.










# Modelling


```{r}
 set.seed(27)

X_train_split <- initial_split(df_train, prop = 0.7)

X_train <- training(X_train_split)

X_test <- testing(X_train_split)



```


```{r}
#model1 <- glm(Survived ~ Sex + Pclass + Fare + Family_size + port_of_embarkment,
           #   data = X_train, family = binomial(link = "logit"))

```


```{r}

# using random forest to fit model

X_train_rf <- X_train[c("Sex", "Family_size", "Pclass", "Fare")]
X_test_rf <- X_test[c("Sex", "Family_size", "Pclass", "Fare")]
y_train <- as.factor(X_train$Survived)

set.seed(27)
rf <- randomForest(x = X_train_rf, y = y_train, importance = TRUE, ntree = 1000)

rf

```


```{r}
predictions <- predict(rf, X_test_rf)

table(predictions)
```

```{r}
library(caret)

confusion <- confusionMatrix(predictions, as.factor(X_test$Survived))
```


```{r}
print(confusion)
accuracy <- confusion$overall["Accuracy"]
print(paste("Accuracy:", accuracy))
```


```{r}

submission <- predict(rf, df_test)

table(submission)



```


```{r}
final_submission <- data.frame(PassengerId = rep(1:417), Survived = submission)


```

```{r}
write.csv(final_submission, file = "RF_Prediction.csv")
```

```{r}

```

```{r}



